Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-09-16 intro.tex
start writing IR divergences

    $f_{a/h}(x, \mu)$ are parton distribution functions
    (PDFs) which depend on the momenta fractions $x$ parton $a$
    has with respect to its parent hadron $h$, at a scale $\mu$,
    usually taken to be the factorisation scale $\mu_{F}$,
    the interface between soft and hard physics.
    The interpretation of PDFs at leading order are
    as probability distributions. $f_{a/h}(x, \mu)$ is
    the probability of finding parton $a$ within $h$
    carrying a momentum fraction $x$ at the energy scale $\mu$.
    PDFs are non-perturbative objects that encapsulate
    the soft effects of the scattering which occur below energy
    $\mu_{F}$. They are non-perturbative because
    they are determined by fitting to experimental data,
    instead of calculated in perturbation theory.
    Multiple groups release PDF fits (cite groups here).
    In practice, these PDF fits
    are accessed through the LHAPDF interface (cite here).
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations (cite here).

>>>>

    $f_{a/h}(x, \mu)$ are parton distribution functions
    (PDFs) which depend on the momenta fractions $x$ parton $a$
    has with respect to its parent hadron $h$, at a scale $\mu$,
    usually taken to be the factorisation scale $\mu_{F}$,
    the interface between soft and hard physics.
    The interpretation of PDFs at leading order are
    as probability distributions. $f_{a/h}(x, \mu)$ is
    the probability of finding parton $a$ within $h$
    carrying a momentum fraction $x$ at the energy scale $\mu$.
    PDFs are non-perturbative objects that encapsulate
    the soft effects of the scattering which occur below energy
    $\mu_{F}$. They are non-perturbative because
    they are determined by fitting to experimental data,
    instead of calculated in perturbation theory.
    Multiple groups release PDF fits (cite groups here).
    In practice, these PDF fits
    are accessed through the LHAPDF interface (cite here).

    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations (cite here)
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\dfrac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $z$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops (cite here). At leading order (cite here) they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

### 2022-09-16 qcd.tex
start writing IR divergences

\section{Divergent structures}
    We have seen that matrix element computations
    result in evaluating Feynman diagrams. From
    the QCD Feynman rules (ref here), it becomes
    clear that integrating over all momentum modes
    will give rise to singularities. The divergences
    associated with high energy modes are ultraviolet
    (UV) divergences. On the opposite end of the energy
    spectrum, low energy modes also give rise to 
    infrared (IR) divergences.

@@

\subsection{Ultraviolet divergences}
    During intermediate steps of calculations,
    such as trying to compute loop diagrams,
    we have to evaluate integrals of the form
    \begin{equation}\label{eqn:loop_integral}
        \int_{0}^{\Lambda} \dfrac{\mathrm{d}^{4}\ell}{(2\pi)^{4}}\dfrac{1}{\ell^{2}(\ell+p)^{2}} \sim \log{\Lambda} \, ,
    \end{equation}
    where a cutoff scale $\Lambda$ has been introduced to
    capture the divergence as the loop momenta
    $k \rightarrow \infty$. It is clear that the
    integral diverges in this high energy limit,
    hence the name ultraviolet divergence.
    This UV divergence cannot influence the
    experimentally observable quantities at experiments,
    which are finite, because the final observable
    should not depend on the intermediate steps taken
    to get there.
    Infinities associated with UV divergences
    are removed via the process of renormalisation
    which introduces counterterms in the Lagrangian
    to exactly cancel these divergences.

@@

\subsection{Renormalisation}
    With the UV divergence regularised by dimensional
    regularisation, it is now possible to construct the forms
    of the counterterms order-by-order in $\alpha_{s}$ to
    explicitly cancel them.

@@

\subsection{Running of the coupling constant}\label{sec:alpha_running}
    The bare coupling should not depend on the actual
    value of the renormalisation scale,
    \begin{equation}\label{eqn:d_g_bare_d_mu}
        \dfrac{\mathrm{d}g_{\mathrm{bare}}}{\mathrm{d}\mu_{R}} = 0 \, ,
    \end{equation}
    or said another way, the renormalisation process
    is independent of the renormalisation scale.

    The behaviour of $\alpha_{s}$
    is governed by the Callan-Symanzik \cite{Callan:1970yg,Symanzik:1970rt}
    $\beta$-function
    \begin{equation}\label{eqn:beta_fn}
        \mu^{2} \dfrac{\partial \alpha_{s}(\mu^{2})}{\partial \mu^{2}} = \beta(\alpha_{s}) \, ,
    \end{equation}
    where the $\beta$-function can be written as a perturbative
    expansion in $\alpha_{s}$
    \begin{equation}\label{eqn:beta_series}
        -\beta(\alpha_{s}) = \alpha_{s}\sum_{n=0}^{\infty}\left(\dfrac{\alpha_{s}}{4\pi}\right)^{n+1} \beta_{n}\, 
    \end{equation}
    where the coefficients $\beta_{n}$ have been computed
    up to fourth order \cite{Baikov:2016tgj,Luthe:2017ttg}.
    The solution to (\ref{eqn:beta_fn}) to first order
    is
    \begin{equation}\label{eqn:1l_alpha}
        \alpha_{s}(\mu^{2}) = \dfrac{1}{\frac{\beta_{0}}{4\pi}\log\left(\frac{\mu^{2}}{\Lambda_{\mathrm{QCD}}^{2}}\right)} \, ,
    \end{equation}
    where $\Lambda_{\mathrm{QCD}}$ is the QCD scale,
    the scale at which $\alpha_{s}$ becomes
    large enough that perturbation theory breaks down,
    and $\beta_{0} = (11C_{A} - \frac{4}{3}T_{R}n_{f})/6\pi$.
    In QCD, $C_{A}=3$ and $T_{R}=1/2$, meaning for
    $n_{f} < 33/2$ the sign of $\beta_{0}$ is positive,
    which is what we have observed in nature with six
    flavours of quarks. This behaviour of positive
    $\beta_{n}$ is also the case in all coefficients
    computed to date, meaning that $\alpha_{s}$
    decreases with increasing energy due to the minus sign
    in (\ref{eqn:beta_series}). This property is known as
    asymptotic freedom \cite{Gross:1973id,Politzer:1973fx},
    and is justification for treating QCD perturbatively.
    The running of the coupling constant has been
    observed experimentally as illustrated in Figure~\ref{fig:alpha_s_running}.
    \begin{figure}
        \includegraphics{qcd/alpha_s_running.pdf}
        \caption{The running of the strong coupling constant $\alpha_{s}$,
        as determined by experiments, with QCD theory prediction in black.
        Figure reference \cite{Workman:2022ynf}.}
        \label{fig:alpha_s_running}
    \end{figure}

    Another consequence of the running coupling
    is colour confinement, in which quarks and
    gluons cannot exist as free particles. Due
    to the increasing strength of coupling at low
    energies, quarks and gluons are forced to
    form composite, colourless particles. This means
    that at low energies, perturbative QCD
    is not an accurate description of nature,
    and the physics of hadrons cannot be predicted
    from perturbation theory.
\subsection{Infrared divergences}
\section{Factorisation of matrix elements}
\subsection{Soft limits}
\subsection{Collinear limits}

>>>>

\section{Divergent structures}\label{sec:divergences}
    We have seen that matrix element computations
    result in evaluating Feynman diagrams. From
    the QCD Feynman rules (ref here), it becomes
    clear that integrating over all momentum modes
    will give rise to singularities. The divergences
    associated with high energy modes are ultraviolet
    (UV) divergences. On the opposite end of the energy
    spectrum, low energy modes also give rise to 
    infrared (IR) divergences.

@@

\subsection{Ultraviolet divergences}\label{sec:UV}
    During intermediate steps of calculations,
    such as trying to compute loop diagrams,
    we have to evaluate integrals of the form
    \begin{equation}\label{eqn:loop_integral}
        \int_{0}^{\Lambda} \dfrac{\mathrm{d}^{4}\ell}{(2\pi)^{4}}\dfrac{1}{\ell^{2}(\ell+p)^{2}} \sim \log{\Lambda} \, ,
    \end{equation}
    where a cutoff scale $\Lambda$ has been introduced to
    capture the divergence as the loop momenta
    $k \rightarrow \infty$. It is clear that the
    integral diverges in this high energy limit,
    hence the name ultraviolet divergence.
    This UV divergence cannot influence the
    experimentally observable quantities at experiments,
    which are finite, because the final observable
    should not depend on the intermediate steps taken
    to get there.
    Infinities associated with UV divergences
    are removed via the process of renormalisation
    which introduces counterterms in the Lagrangian
    to exactly cancel these divergences.

@@

\subsection{Renormalisation}\label{sec:renormalisation}
    With the UV divergence regularised by dimensional
    regularisation, it is now possible to construct the forms
    of the counterterms order-by-order in $\alpha_{s}$ to
    explicitly cancel them.

@@

\subsection{Running of the coupling constant}\label{sec:alpha_running}
    The bare coupling should not depend on the actual
    value of the renormalisation scale,
    \begin{equation}\label{eqn:d_g_bare_d_mu}
        \dfrac{\mathrm{d}g_{\mathrm{bare}}}{\mathrm{d}\mu_{R}} = 0 \, ,
    \end{equation}
    or said another way, the renormalisation process
    is independent of the renormalisation scale. The
    consequence of this is that $\alpha_{s}$ has to
    depends on the renormalisation scale.

    The behaviour of $\alpha_{s}$
    is governed by the Callan-Symanzik \cite{Callan:1970yg,Symanzik:1970rt}
    $\beta$-function
    \begin{equation}\label{eqn:beta_fn}
        \mu^{2} \dfrac{\partial \alpha_{s}(\mu^{2})}{\partial \mu^{2}} = \beta(\alpha_{s}) \, ,
    \end{equation}
    where the $\beta$-function can be written as a perturbative
    expansion in $\alpha_{s}$
    \begin{equation}\label{eqn:beta_series}
        -\beta(\alpha_{s}) = \alpha_{s}\sum_{n=0}^{\infty}\left(\dfrac{\alpha_{s}}{4\pi}\right)^{n+1} \beta_{n}\, 
    \end{equation}
    where the coefficients $\beta_{n}$ have been computed
    up to $\beta_{4}$ \cite{Baikov:2016tgj,Luthe:2017ttg}.
    The solution to (\ref{eqn:beta_fn}) to first order
    is
    \begin{equation}\label{eqn:1l_alpha}
        \alpha_{s}(\mu^{2}) = \dfrac{1}{\frac{\beta_{0}}{4\pi}\log\left(\frac{\mu^{2}}{\Lambda_{\mathrm{QCD}}^{2}}\right)} \, ,
    \end{equation}
    where $\Lambda_{\mathrm{QCD}}$ is the QCD scale,
    the scale at which $\alpha_{s}$ becomes
    large enough that perturbation theory breaks down,
    and $\beta_{0} = (11C_{A} - 4T_{R}n_{f})/3$.
    In QCD, $C_{A}=3$ and $T_{R}=\frac{1}{2}$, meaning for
    $n_{f} < \frac{33}{2}$ the sign of $\beta_{0}$ is positive,
    which is what we have observed in nature with six
    flavours of quarks. This behaviour of positive
    $\beta_{n}$ is also the case in all coefficients
    computed to date, meaning that $\alpha_{s}$
    decreases with increasing energy due to the minus sign
    in (\ref{eqn:beta_series}). This property is known as
    asymptotic freedom \cite{Gross:1973id,Politzer:1973fx},
    and is justification for treating QCD perturbatively.
    The running of the coupling constant has been
    observed experimentally as illustrated in Figure~\ref{fig:alpha_s_running}.
    \begin{figure}
        \includegraphics{qcd/alpha_s_running.pdf}
        \caption{The running of the strong coupling constant $\alpha_{s}$,
        as determined by experiments, with QCD theory prediction in black.
        Figure reference \cite{Workman:2022ynf}.}
        \label{fig:alpha_s_running}
    \end{figure}

    Another consequence of the running coupling
    is colour confinement, in which quarks and
    gluons cannot exist as free particles. Due
    to the increasing strength of coupling at low
    energies, quarks and gluons are forced to
    form composite, colourless particles. This means
    that at low energies, perturbative QCD
    is not an accurate description of nature,
    and the behaviour of hadrons cannot be predicted
    from perturbation theory.
\subsection{Infrared divergences}
    In Section~\ref{sec:UV} we saw how divergences
    stemming from high energy behaviour would occur
    in loop diagrams. On the other end of the energy
    spectrum we also have divergences from low energy
    modes. Since they are related to low energy
    behaviours they are called IR divergences. There
    are two cases in which these divergences arise:
    in loop integrals (virtual) and when an emission
    of an extra particle becomes soft or collinear
    (real).

    Virtual divergences can be understood by looking
    at the integral (\ref{eqn:loop_integral})

\section{Factorisation of matrix elements}
\subsection{Soft limits}
\subsection{Collinear limits}
